A High-Level Embedded Language for Low-Level GPU Kernel Programming

### 2013-11-03 src2/hl/paperThesis.tex
update

Our work on Obsidian makes the following contributions: 
\begin{itemize} 
  \item A high-level programming interface that encourages design space exploration through experimentation with 
    details of GPU kernels (section \ref{sec:CASESTUDIES}).
  \item Compositional abstractions for implementation of efficient GPU 
    Kernels (section \ref{sec:CASESTUDIES}).
  \item We use types to model the hierarchical nature of GPUs in our embedded 
    language and to rule out programs that we cannot easily compile to a GPU (section \ref{sec:Program}).
  \item We use type directed compilation of language constructs and 
    implement some GPU hierarchy generic functions (section \ref{sec:Compile}).
  \item A simple and compositional method for reification of monads in the 
    setting of a compiled embedded language (section \ref{sec:Compile}).

\end{itemize}

@@

\end{verbatim}
\end{small}
\caption{Code generated from the {\tt reverses} program with the local reverse instantiated with 
{\tt reverseL} and {\tt reverseL'}. 
 {\tt reverse'} stores the array in shared memory before copying it to the global result array.} 
\label{fig:rev1} 
\end{figure}

@@

The following Obsidian code shows how to split a large array up into 
chunks of 256 elements ({\tt splitUp 256 arr}) and how the local reverse 
kernel is mapped, in parallel, over blocks using {\tt pConcatMap}.

\begin{small} 
\begin{verbatim} 
reverses :: DPull a -> DPush Grid a
reverses arr = pConcatMap reverseL $ splitUp 256 arr 
\end{verbatim} 
\end{small}

\noindent
The {\tt pConcatMap} combinator has the following type: 
%takes a local computation that results in a push array.
\begin{small}
\begin{Verbatim}[samepage=true] 
pConcatMap :: ASize l 
              => (a1 -> Program t (SPush t a)) 
              -> Pull l a1 
              -> Push (Step t) l a
\end{Verbatim}
\end{small}

{\tt pConcatMap} takes a computation resulting in a push array at a specific 
level of the GPU hierarchy, {\tt t}, and a pull array. The computation 
is applied to each element of the pull array and the results are concatenated 
into a push array at the level above {\tt t}. Here, this is used to apply a block 
level computation, giving a grid computation.

@@

\begin{small} 
\begin{verbatim} 
largeReverse :: DPull a
                -> Push Grid EWord32 a
largeReverse arr =
  pConcat $ reverse $ pMap reverseL $ splitUp 256 arr 
\end{verbatim} 
\end{small}

@@

Very similar representations are used 
in many other embedded languages. The 
embedded language Pan ~\citehl{COMPILEEDSL} used a similar representation for 
images and it has been used since in Obsidian, Feldspar ~\citehl{FELDSPAR} 
and Repa ~\citehl{REPA}. In Obsidian, there is a {\tt force} function that,
when applied to a pull array, writes the elements of that array 
into shared memory using as many threads as the pull array is long. Hence, 
the {\tt force} function can be used to express parallelism in Obsidian.

@@

The length of an array, the {\tt s} parameter,
can be either static (a Haskell known-at-compile-time value) or dynamic 
(a runtime value). Static lengths are used for local (or block) 
computations, with those lengths determining the number of threads needed and the shared memory consumption. Dynamic 
lengths are used for global computations (across many blocks). After specifying 
a local computation, it can often be applied over a variable number of blocks. 
The dynamic lengths capture this common use case.

@@

Both push and pull arrays represent methods of computing an array. They 
do not directly correspond to actual data stored in memory, but rather 
are virtual. This enables fusion of operations on both push and pull arrays. 
One such example is map fusion: \verb!map f . map g = map (f . g)!. It is 
only by using {\tt force} that the programmer can explicitly demand that the 
operations should not be fused but that the actual intermediate result should 
be stored in memory. In Obsidian, {\tt force} is also the only way 
to convert a push array into a pull array. 
The type of {\tt force} is: 
\begin{small} 
\begin{Verbatim}[samepage=true] 
force :: (Write p, MemoryOps a) 
         => p Word32 a 
         -> Program Block (Pull Word32 a)
\end{Verbatim}
\end{small}

\noindent
where {\tt p} is an array (push or pull) with elements that can be stored into memory. 
Note that the return type of {\tt force} is a program yielding a pull array. This pull 
array represents reading from a named area of memory.

@@

After compiling the monadic {\tt Program} into a list of statements, 
some characteristics are collected.

The first analysis done on the {\tt IM} 
discovers the number of threads needed per block to run the program. This 
is done by a pass over the {\tt IM} and looking at the size  
of each parallel {\tt SForAll} loop. The largest {\tt SForAll} loop present 
decides how many threads are needed. The implementation of any other 
{\tt SForAll} loop will be wrapped in a conditional that turns 
off any unused threads.

Note that there is no similar analysis to discover the maximum extents 
of {\tt SForAllBlocks} loops. The reason is that the restriction
to kernels of the form:

\noindent means that there can be no sequence of {\tt SForAllBlocks} loops in the program. In other 
words, an Obsidian Program for which we can generate code will contain at most 
one {\tt SForAllBlocks} loop. Some care has also to be taken with
the language constructs that result in grid push arrays, so that none of these 
creates illegal programs.

@@

In the current {\tt IM}, all arrays in shared memory will have unique 
names. These names are obtained via the value supply during the reification 
of the {\tt Program} monad. These names will be used in a memory mapping phase.
A {\em liveness} analysis of the arrays collects information needed to lay 
out the arrays in the shared memory of the GPU. Currently no spilling of 
arrays that do not fit into shared memory is performed, so it is very possible 
to design an Obsidian program that will not work because of shared memory 
restrictions. After the arrays are assigned addresses in the shared memory,
a renaming phase replaces array names with these locations. This is done 
while also going from the {\tt IM} data type to a third internal representation 
of the program. This third representation is a C-like AST extended with some 
GPU specific concepts (just as CUDA is an extended C). We strive to make this representation sufficiently general to be turned into 
either CUDA code or OpenCL code, but currently only output CUDA.

@@

%% \begin{small} 
%% \begin{verbatim} 
%% iters bid tid = 
%%   liftM extract $ seqUntil (f bid tid) cond  (0,0,1)
%%   where
%%     extract (_,_,c) = ((word32ToWord8 c) `mod` 16) * 16
%% \end{verbatim}
%% \end{small}
The final step is to run the iterations for each pixel location. 
This is done by using the {\tt generate} function: 
\begin{small}
\begin{Verbatim}[samepage=true] 
generate :: ASize l
            => l
            -> (EWord32 -> Program t (SPush t b))
            -> Push (Step t)  l b
\end{Verbatim}
\end{small}

@@

\begin{small} 
\begin{verbatim} 
genRect :: MemoryOps b
           => EWord32
           -> Word32
           -> (EWord32 -> EWord32 
                 -> TProgram (SPush Thread b))
           -> DPush Grid b 
genRect bs ts p = 
   generate bs (\bid -> return $ generate ts (p bid))
\end{verbatim}
\end{small} %$

The largest possible image that can be obtained using this particular 
method is 1024$\times$1024. This is because of the maximum threads per block 
limitation of current GPUs. The code below generates a 512$\times$512 element 
image. 
\begin{small} 
\begin{verbatim}
mandel = genRect 512 512 iters 
\end{verbatim}
\end{small}

@@

\begin{small} 
\begin{Verbatim}[samepage = true] 
mapRed1 :: MemoryOps a 
              => (a -> a -> a) 
              -> DPull (SPull a) 
              -> DPush Grid a
mapRed1 f = pConcatMap $ red1 f
\end{Verbatim}
\end{small} %$

>>>>

Our work on Obsidian makes the following contributions: 
\begin{itemize} 
  \item A high-level programming interface that encourages design space exploration through experimentation with 
    details of GPU kernels (section \ref{sec:CASESTUDIES}).
  \item Compositional abstractions for implementation of efficient GPU 
    Kernels (section \ref{sec:CASESTUDIES}).
  \item We use types to model the hierarchical nature of GPUs in our embedded 
    language and to rule out programs that we cannot easily compile to a GPU (section \ref{sec:Program}).
  \item We use type directed compilation of language constructs and 
    implement some GPU hierarchy generic functions (section \ref{sec:Compile}).
 % \item A simple and compositional method for reification of monads in the 
 %   setting of a compiled embedded language (section \ref{sec:Compile}).
\end{itemize}

@@

\end{verbatim}
\end{small}
\caption{Code generated from the {\tt reverses} and {\tt reverses'} programs.
 {\tt reverses'} stores the array in shared memory before copying it to the global result array.} 
\label{fig:rev1} 
\end{figure}

@@

The following Obsidian code shows how to split a large array up into 
chunks of 256 elements ({\tt splitUp 256 arr}) and how the local {\tt reverseL} 
and {\tt reverseL'} kernels are  mapped, in parallel, over blocks using {\tt pConcatMap}.

\begin{small} 
\begin{verbatim} 
reverses :: DPull a -> DPush Grid a
reverses arr = pConcatMap (pJoin . reverseLocal)
                          (splitUp 256 arr)

reverses' :: MemoryOps a => DPull a -> DPush Grid a
reverses' arr = pConcatMap (pJoin . reverseLocal')
                           (splitUp 256 arr) 
\end{verbatim} 
\end{small}

\noindent
The {\tt pConcatMap} combinator has the following type: 
%takes a local computation that results in a push array.
\begin{small}
\begin{Verbatim}[samepage=true] 
pConcatMap :: ASize l 
            => (a1 -> SPush t a) 
            -> Pull l a1 
            -> Push (Step t) l a
\end{Verbatim}
\end{small}

{\tt pConcatMap} takes a computation resulting in a push array at a specific 
level of the GPU hierarchy, {\tt t}, and a pull array. The computation 
is applied to each element of the pull array and the results are concatenated 
into a push array at the level above {\tt t}. Here, this is used to apply a block 
level computation, giving a grid computation. The {\tt pJoin} function 
has type \verb!Program t (Push t s a) -> Push t s a!; it creates a new push array 
that computes the same results as the input computation.

@@

\begin{small} 
\begin{verbatim} 
largeReverse :: DPull b
              -> Push Grid EWord32 b
largeReverse arr =
  pConcat $ reverse $ fmap (pJoin . reverseLocal)
                      (splitUp 256 arr) 
\end{verbatim} 
\end{small}

@@

Very similar representations are used 
in many other embedded languages. The 
embedded language Pan ~\citehl{COMPILEEDSL} used a similar representation for 
images and it has been used since in Obsidian, Feldspar ~\citehl{FELDSPAR} 
and Repa ~\citehl{REPA}. In Obsidian, there is a {\tt force} function that,
when applied to a pull array, writes the elements of that array 
into shared memory using as many threads as needed from those available. 
If the array is larger than the number of threads made available, the 
force operation is split into two or more passes. This is a change compared
to earlier versions of Obsidian where the length of arrays we could force 
where limited to the maximum number of threads. In this version, the 
threads are, in a sense, virtual. The mapping from virtual threads to 
real threads is done as a step in the compilation.

@@

The length of an array, the {\tt s} parameter,
can be either static (a Haskell known-at-compile-time value) or dynamic 
(a runtime value). Static lengths are used for local (or block) 
computations, with those lengths determining shared memory consumption. Dynamic 
lengths are used for global computations (across many blocks). After specifying 
a local computation, it can often be applied over a variable number of blocks. 
The dynamic lengths capture this common use case.

@@

Both push and pull arrays represent methods of computing an array. They 
do not directly correspond to actual data stored in memory, but rather 
are virtual. This enables fusion of operations on both push and pull arrays. 
One such example is map fusion: \verb!map f . map g = map (f . g)!. It is 
only by using {\tt force} that the programmer can explicitly demand that the 
operations should not be fused but that the actual intermediate result should 
be stored in memory. In Obsidian, {\tt force} is also the only way 
to convert a push array into a pull array. 
With some class constraints removed for simplicity, the type of {\tt force} is: 
\begin{small} 
\begin{Verbatim}[samepage=true] 
force :: ( ... ) 
       => arr Word32 a 
       -> Program Block (Pull Word32 a)
\end{Verbatim}
\end{small}

\noindent
where {\tt arr} is an array (push or pull) with elements that can be stored into memory. 
Note that the return type of {\tt force} is a program yielding a pull array. This pull 
array represents reading from a named area of memory.

@@

%% After compiling the monadic {\tt Program} into a list of statements, 
%% some characteristics are collected.

%% The first analysis done on the {\tt IM} 
%% discovers the number of threads needed per block to run the program. This 
%% is done by a pass over the {\tt IM} and looking at the size  
%% of each parallel {\tt SForAll} loop. The largest {\tt SForAll} loop present 
%% decides how many threads are needed. The implementation of any other 
%% {\tt SForAll} loop will be wrapped in a conditional that turns 
%% off any unused threads.

%% Note that there is no similar analysis to discover the maximum extents 
%% of {\tt SForAllBlocks} loops. The reason is that the restriction
%% to kernels of the form:

%% \begin{small} 
%% \begin{Verbatim}[samepage=true]
%% myKernel :: a0 -> ... -> aN -> DPush Grid b 
%% \end{Verbatim} 
%% \end{small}

%% \noindent means that there can be no sequence of {\tt SForAllBlocks} loops in the program. In other 
%% words, an Obsidian Program for which we can generate code will contain at most 
%% one {\tt SForAllBlocks} loop. Some care has also to be taken with
%% the language constructs that result in grid push arrays, so that none of these 
%% creates illegal programs.

@@

In the {\tt IM}, all arrays in shared memory will have unique 
names. These names are obtained via the value supply during the reification 
of the {\tt Program} monad. These names will be used in a memory mapping phase.
A {\em liveness} analysis of the arrays collects information needed to lay 
out the arrays in the shared memory of the GPU. Currently no spilling of 
arrays that do not fit into shared memory is performed, so it is very possible 
to design an Obsidian program that will not work because of shared memory 
restrictions. After the arrays are assigned addresses in the shared memory,
a renaming phase replaces array names with these locations.

The last step in compilation process uses the Language.C.Quote library to 
turn the {\tt IM} into CUDA code. This compilation stage takes a {\em configuration} 
as parameter. The configuration tells how many actual threads per block the 
CUDA code should be generated for. At this stage parallel for loops (\tt SForAll) 
are turned into more than one parallel for loop in sequence, if needed.

%This is done 
%while also going from the {\tt IM} data type to a third internal representation 
%of the program. This third representation is a C-like AST extended with some 
%GPU specific concepts (just as CUDA is an extended C). We strive to make this representation sufficiently general to be turned into 
%either CUDA code or OpenCL code, but currently only output CUDA.

@@

%% \begin{small} 
%% \begin{verbatim} 
%% iters bid tid = 
%%   liftM extract $ seqUntil (f bid tid) cond  (0,0,1)
%%   where
%%     extract (_,_,c) = ((word32ToWord8 c) `mod` 16) * 16
%% \end{verbatim}
%% \end{small}
The final step is to run the iterations for each pixel location. 
This is done by using the {\tt generate} function: 
\begin{small}
\begin{Verbatim}[samepage=true] 
generate :: ASize l 
          => l 
          -> (EWord32 -> SPush t b) 
          -> Push (Step t) l b
\end{Verbatim}
\end{small}

@@

\begin{small} 
\begin{verbatim} 
genRect :: MemoryOps b
           => EWord32
           -> Word32
           -> (EWord32 -> EWord32 -> SPush Thread b)
           -> DPush Grid b 
genRect bs ts p = generate bs $
                  \bid -> generate ts $ p bid 
\end{verbatim}
\end{small} %$

The largest possible image that can be obtained using this particular 
method is 1024$\times$1024. This is because of the maximum threads per block 
limitation of current GPUs. The code below generates a 512$\times$512 element 
image. 
\begin{small} 
\begin{verbatim}
mandel = genRect 512 512 (\b -> pJoin . iters b)
\end{verbatim}
\end{small}

@@

\begin{small} 
\begin{Verbatim}[samepage = true] 
mapRed1 :: MemoryOps a 
              => (a -> a -> a) 
              -> DPull (SPull a) 
              -> DPush Grid a
mapRed1 f = pConcatMap $ pJoin . red1 f
\end{Verbatim}
\end{small} %$
